### 2024-04-30 paper.tex
two minor changes
maybe we dont need 3 references for seurat :D

Lukas P.~M. Kremer\textsuperscript{1,2,*},
Martina M. Braun\textsuperscript{2},
Svetlana Ovchinnikova\textsuperscript{1},
Leonie Küchenhoff\textsuperscript{1,2},\\
Santiago Cerrizuela\textsuperscript{2},
Ana Martin-Villalba\textsuperscript{2,*},
Simon Anders\textsuperscript{1,*}\\[2em]
}
\noindent\textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\
\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany\\
\textsuperscript{*} e-mails:\\
\texttt{l.kremer@dkfz-heidelberg.de}\\
\texttt{a.martin-villalba@dkfz-heidelberg.de}\\
\texttt{simon.anders@bioquant.uni-heidelberg.de}

@@

The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.\ counts of distinct RNA molecules), one row for each cell and one column for each gene.
A first goal is usually to assign cell types or states to cells.
To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or Scanpy \citep{Wolf_2018}, is as follows:
One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts.
Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes.
Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

LPMK and SA conceived and worked out the method.
LPKM wrote the software package, with contributions from MMB (differential methylation functionality) and LKü.
LPMK, SO and LKü prepared the use-case demonstrations and performed the benchmarks.
SC and AMV contributed experimental data and biological expertise.
LPMK, AMV and SA contributed supervision and project management.
LPMK and SA wrote the manuscript.

@@

%\phantomsection
\clearpage
\section*{Methods}

@@

\subsection*{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

>>>>

Lukas P.~M. Kremer\textsuperscript{1,2,*},
Martina M. Braun\textsuperscript{2},
Svetlana Ovchinnikova\textsuperscript{1},
Leonie Küchenhoff\textsuperscript{1,2},\\
Santiago Cerrizuela\textsuperscript{2},
Ana Martin-Villalba\textsuperscript{2,*},
Simon Anders\textsuperscript{1,*}\\[2em]
}
\noindent\textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\
\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany\\
\textsuperscript{*} corresponding authors. e-mails:\\
\texttt{l.kremer@dkfz-heidelberg.de}\\
\texttt{a.martin-villalba@dkfz-heidelberg.de}\\
\texttt{simon.anders@bioquant.uni-heidelberg.de}

@@

The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.\ counts of distinct RNA molecules), one row for each cell and one column for each gene.
A first goal is usually to assign cell types or states to cells.
To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{seurat5} or Scanpy \citep{Wolf_2018}, is as follows:
One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts.
Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes.
Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

LPMK and SA conceived and worked out the method.
LPKM wrote the software package, with contributions from MMB (differential methylation functionality) and LKü.
LPMK, SO and LKü prepared the use-case demonstrations and performed the benchmarks.
SC and AMV contributed experimental data and biological expertise.
LPMK, AMV and SA contributed supervision and project management.
LPMK and SA wrote the paper.

@@

\clearpage
\section*{Methods}

@@

\subsection*{Mean neighbor score} \label{methods:score}
To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat5} function \texttt{FindClusters(resolution = 0.1)}.
The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded.
The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
